---
title: "Lab 2"
subtitle: "Introductory Computing -2"
author: "Qing Zhang"
date: "Updated: `r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false


---


# This week:
## Introductory Computing 2

- Odyssey Intro
- Odyssey Demo
  - Login
  - File transfer
  - Navigation, HW7
  - For-loop
  - sbatch script
- BWA and STAR
- HW7 potential gene
- Dynamic programming


---
# Odyssey

- What is Odyssey? Lots of computers stringed together
- Advantage: More storage, can run many things in parallel (e.g. use 10 computers to process 10 samples at a time)
- Disadvantage: a lot of overhead to get things to work (have to make sure your stuff doesn't interfere with other people's stuff)
- Can't just run stuff through the terminal on Odyssey, you will need to submit job using srun or sbatch (preferred)
- My tip: start off by requesting very few resources and doing a test run on a small file.
- **START EARLY**. You might have to wait a long time for your jobs to run.
- Odyssey might also be slow if everyone in the class tries at
the last minute.

---
## Odyssey Logistics

- Login using ssh (Mac/Linux) or PuTTY (Windows)
- Need both your password and 6-digit code, either on your phone or
as a Java applet on your computer.
- Transfer files using Filezilla or WinSCP.
- Details: [https://www.rc.fas.harvard.edu/resources/odyssey-quickstart-guide/](https://www.rc.fas.harvard.edu/resources/odyssey-quickstart-guide/)


---

## Basic Bash Commands

- `pwd`: Print Working Directory. Where am I?
- `cd folder`: Change directories.
- `cd`: Shortcut to go back to home directory.
- `..`: Shortcut for going up a directory.
    - `cd ..`
    - `cd ../../`
    - `cd ../folder`
- `~`: shortcut for home directory. Can use it like so: `cd ~/stat115-hw2`.
- `mkdir folder`: Make a directory called `folder`.
- `less file`: View a file. Press `q` to quit.
- `man cmd`: Get help on `cmd`. Press `q` to quit.
- Videos about command line: [https://www.youtube.com/watch?v=U3iNcBtycaQ](https://www.youtube.com/watch?v=U3iNcBtycaQ)

---

# Odyssey Commands

Odyssey uses **slurm** workload manager.

- Submit a batch serial job: 	`sbatch runscript.sh`
- Run a script or application interactively: `srun --pty -p test -t 10 --mem 1000 /bin/bash [script or app]`
- Start interactive session: `srun --pty -p test -t 10 --mem 1000 /bin/bash`
- Kill a job: `scancel 999999`
- View status of your jobs :`squeue -u username`
- Check current job by id number: `sacct -j 999999`


- Can check job status online, too: [https://portal.rc.fas.harvard.edu/login/?next=/jobs/](https://portal.rc.fas.harvard.edu/login/?next=/jobs/)



---

## Example Submission Script

- Save this to a file in your home directory (e.g. `submit.sh`).
    + Good idea to make a `stat115-hw2` folder
- Submit by running `sbatch submit.sh`.
- Can edit on your own computer using a a plain text editor (e.g.
Sublime or TextEdit)
- Transfer the file using Filezilla/WinSCP.

```{r engine = "bash", eval = FALSE}
#!/bin/bash
#SBATCH -n 1 # Number of cores requested
#SBATCH -N 1 # Ensure that all cores are on one machine
#SBATCH -t 15 # Runtime in minutes
#SBATCH -p serial_requeue # Partition to submit to
#SBATCH --mem=1G # Memory in GB (see also --mem-per-cpu)
#SBATCH -o output_%j.out # Standard out goes to this file
#SBATCH -e error_%j.err # Standard err goes to this file

# LOAD_MODULES
module load bwa/0.7.15-fasrc02


# YOUR_COMMANDS_HERE
echo JOB_FINISHED
```


---

## Submission Script Recommendations

- Runtime: 30 min for BWA, 20 min for STAR
- Memory: 5G for BWA, 32G for STAR
- Change the name of the `.out` and `.err` files to something more
meaningful.
- Permission: You cannot write to class folder - please read files from class folder and save output to your somewhere under your home directory.
- Stick with `serial_requeue`. You'll have to wait less.
- Need to load modules for the tools we want to use.
- Modules for BWA:
``` 
module load bwa/0.7.15-fasrc02
module load samtools/1.5-fasrc02
```
- Modules for STAR:
```
module load STAR/2.6.0c-fasrc01
```
- Module search: [https://portal.rc.fas.harvard.edu/p3/build-reports/](https://portal.rc.fas.harvard.edu/p3/build-reports/)


---

# Odyssey Demo


---

# BWA

- Uses Burrows-Wheeler transform.
- Requires an index to be made before aligning.
- 3 different versions:
    + BWA-backtrack (`aln/samse/sampe`)
    + BWA-SW (`bwasw`)
    + BWA-MEM (`mem`)
- Different recommendations depending on data:
    + BWA-backtrack: Illumina reads up to 100bp
    + BWA-SW / BWA-MEM: longer sequeces from 70bp to 1Mbp
    + BWA-MEM latest version, recommended since generally faster and
    more accurate
    + BWA-MEM better than BWA-backtrack for 70-100bp Illumina reads
- We have short reads, so for the HW, we recommend `bwa aln`.



---
- How to run (basic):
    - Replace `ref.fasta` and `read1.fastq` with the path on Odyssey.

```{r engine = "bash", eval = FALSE}
# align each end separately
bwa aln ref.fasta read1.fastq > read1.sai
bwa aln ref.fasta read2.fastq > read2.sai

# stick everything in a SAM file
bwa sampe ref.fasta read1.sai read2.sai read1.fastq read2.fastq > aln-pe.sam

# count number of aligned reads, etc.
samtools flagstat aln-pe.sam

# count number of uniquely mapped reads
samtools view -h aln_pe.sam | grep -v -e 'XA:Z:'  -e 'SA:Z:'| samtools view -F 0x900 -f 0x2 -q 5| wc -l
```


---
# STAR

- [Star: ultrafast universal RNA-seq aligner](https://academic.oup.com/bioinformatics/article/29/1/15/272537)
    - Suffix array (SA) approach to leverage speed of binary searching.
    - Requires an index (the SA) to be created prior to aligning.
- Splice-aware aligner.
- Replace STARIndex with path of STARIndex.
- Replace `read1.fastq` and `read2.fastq` with path to the reads.
- Replace `output` with something more meaningful.

```{r, engine = "bash", eval = FALSE}
STAR --genomeDir STARIndex \
--readFilesIn read1.fastq read2.fastq \
--runThreadN 1 --outFileNamePrefix output 
```

---
# HW1- Question 7

a. How many lines are there in this file? How many times "KRAS" gene 
has appeared?

```
command to subset | count the lines
```

- `grep`
- `wc`

---

b. How many unique `Variant_Classification` are there in the MAF? Please 
count occurence of each type and sort them. Which one is the most frequent? 

c. What are the top FIVE most frequent genes? Please provide 
the bash command and equivalent Python command. If you are a PI 
looking for a gene to investigate (you need to find a gene with potentially 
better biological significance), which gene out of the top 5 would you 
choose? Why?

**Similar question: how to get value_counts for a column?**

```
 cut -f${column_num} ${tab_path} | sort | uniq -c | sort -n
```

---

## Equivalent Python Command

- Read file and save the colunm as a list

```
with open('/n/stat115/2020/HW1/public_MC3.maf') as f:
    lines = [line.rstrip().split('\t')[2] for line in f]
```

- Count values from a list

```
from collections import Counter
cc = Counter(lines)
```

- Find the top N frequent genes (get the indices)

```
sorted(range(len(cc.values())), key=lambda i: cc.values()[i])[-N:]
```

---

# How to choose a gene of biological significance?

- Cancer is "uncontrolled cell division"" and is caused by mutation. 
- We are interested in genes with *recurrent* mutations that occurs on many patients over different cancer types.
  - However, mutations are usually malignant (due to evolution)
  - The ones that are persisent usually give cells some advantage (e.g. to proliferate more, escape proofread mechanism in replication).

----
We can repharse the question as:
> Which gene is significantly more likely to mutate than by chance?

- Literature is great!
- Null hypothesis?
  - Assume each base is equally mutable:
    - What will happen to long gene?
    - What if slience mutation dominates?
    - Let checking mutation landscape using [Tumor Portal](http://www.tumorportal.org/)


---

# Dynamic programming

- Find the maximum subset sum. Print out both the sum and the starting
and ending index.
- Must use dynamic programming: only loop through the array once.
- Dynamic programming: solve overlapping sub-problems.
- Notation: array $A$ with $k$ elements $\{A_0, A_1, \ldots, A_{k - 1}\}$.
- If we know the maximum subset sum up to position i is $B_i$, what is
the maximum sum up to position $i + 1$?
    - $B_{i + 1} = \max(B_i + A_{i + 1}, A_{i+1})$.
    - Can loop through the array and iterate over $i$.
    - Keep track of the maximum of the $B_i$.
    - Maximum subset sum is $\max_{i} B_i$.

---

### Problem 9: Indexes, runtime, and memory.

- End index of maximum subset sum is the maximizing $i$.
- Start index:
    - Start index for a sequence changes when we pick $A_{i + 1}$
    instead of $A_{i + 1} + B_{i}$.
    - Keep track of the current start index.
    - When the maximum $B_i$ changes, update the start index of the
    maximum subset sum.
- Only need to loop through the array once. Only 1 `for` loop.
- Memory: Don't need to store all the $B_i$, just the previous one.
- $O(1)$ space and $O(n)$ runtime.


---

# Finally...

- Please do not make your homework public on Github.
- Office Hour: Friday 4:30-5:30 PM HSPH Kresge Cafe

